rObsPack: An R package for the NOAA global observations ObsPack methane GLOBALVIEW+

### 2023-04-20 paper/paper.Rmd
first commit

---
title: "robspack"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

![GitHub commit
activity](https://img.shields.io/github/commit-activity/y/ibarraespinosa/robspack)
[![R build
status](https://github.com/ibarraespinosa/robspack/workflows/R-CMD-check/badge.svg)](https://github.com/ibarraespinosa/robspack/actions)
[![R build
status](https://github.com/ibarraespinosa/robspack/workflows/draft-pdf/badge.svg)](https://github.com/ibarraespinosa/robspack/actions)

[NOAA Obspack](https://gml.noaa.gov/ccgg/obspack/) is a collection of green house gases observations

`robspack` only depends on `data.table`, which is basically parallel C, 
so it can be installed in any machine.

## installation

```{r, eval = F}
remotes::install_github("ibarraespinosa/robspack")
```

```{r, include = T, message=F, warning=F}
library(robspack)
library(data.table)
```

@@

categories <- c(
  "aircraft-pfp",
  "aircraft-insitu",
  "surface-insitu",
  "aircore",
  "surface-pfp",
  "tower-insitu",
  "shipboard-insitu",
  "flask"
)
obs <- "../../obspack_ch4_1_GLOBALVIEWplus_v4.0_2021-10-14/data/txt"
index <- obs_summary(obs = obs)

@@

```{r, eval = T}
tmp <- tempfile()
master <- df5
master$timeUTC <- as.character(master$timeUTC)
master$timeUTC_end <- as.character(master$timeUTC_end)
master$local_time <- as.character(master$local_time)

fwrite(master, 
       file =  "paper/tower_insitu_2020.txt",
       sep = " ")
```

The format Comma Separated Value with YAML (CSVY)[^3]
consists in a typical CSV with a YAML header.
The function`obs_write_csvy` includes the argument
`notes` which allows adding custom notes at the header of the file. Below the notes, `obs_write_csvy` adds 
the output of the R function `str`, which
provides a vertical summary of the data, known as
structure.

```{r, eval = T}
obs_write_csvy(dt = master,
          notes = "tower 2020",
          out = "paper/tower_insitu_2020.csvy")
```

@@

```{r, eval = F}
readLines("paper/tower_insitu_2020.csvy")[1:38]
```

@@

## Recommendation for other applications
The approach to generate receptors depends on each type of 
observation and other considerations. For instance, aircraft with 
continuous observations at each second can be filtered and averaged 
every 20 seconds. In that way, the footprints are still 
representative and it would not be necessary to run HYSPLIT 
every second. Of course, it depends on the application 
and objective of the study. For this manuscript, we are 
presenting the generation of receptors based on tower observations.

## Application for other sectors
In this package we are sharing scripts to process other sectors
The scripts are available in the path 
`https://github.com/ibarraespinosa/robspack/tree/main/rscripts`

>>>>

---
# Example from https://joss.readthedocs.io/en/latest/submitting.html
title: 'rObsPack: An R package for the NOAA global observations ObsPack methane GLOBALVIEW+'
tags:
  - R
  - NOAA
  - CH4
  - methane
  - ObsPack
authors:
  - name: Sergio Ibarra-Espinosa
    orcid: 0000-0002-3162-1905
    equal-contrib: true
    affiliation: "1, 2"
  - name: Lei Hu
    orcid: 0000-0002-5624-3832
    equal-contrib: true
    affiliation: "2"
affiliations:
 - name: Cooperative Institute for Research in Environmental Sciences, University of Colorado Boulder, Boulder, CO, USA
   index: 1
 - name: NOAA Global Monitoring Laboratory, Boulder, CO, USA
   index: 2
citation_author: Ibarra-Espinosa and Hu
date: 04 April 2023
year: 2023
bibliography: paper.bib
output: rticles::joss_article
csl: apa.csl
journal: JOSS
---
<!-- output: rticles::joss_article -->
<!-- output: word_document -->

# Summary

Observations of greenhouse gas emissions are critical to monitor 
the state of the atmosphere, quantify present and historical 
emissions, and understand global climate change. Observation 
Package (ObsPack) is a specific format to deliver atmospheric
measurements data of trace gasses [@masarie2014obspack]. 
The NOAA Global Monitoring Laboratory (GML) manages a 
global network of atmospheric observations compiled and 
delivered to the public as ObsPack CH4 GLOBALVIEW+. This 
product provides access to global methane observations made 
in surface, tower, aircrafts, aircores and ships. Processing 
global data can be cumbersome since they are provided as 
multiple text and NetCDF files. The text files consist of a 
commented metadata followed with the tabulated data which can 
have different headers across files. The metadata can have critical
information which determines the nature of the file itself, such as
altitude of the site, local hour, laboratory and more. Below the
metadata, starts the section with tabulated data, with different
header across files.

# Statement of need

In order to analyze greenhouse gasses measurements we developed the 
R package `robspack`, which reads and processes NOAA/GML ObsPack
GLOBALVIEW+ CH4 data. This package imports functions from 
`data.table` R package, which contains C bindings with 
parallel implementation via Open-MP [@dt]. 
`data.table` is faster than other Python, Julia and R 
implementations for data-science, providing a strong basis
for `robspack`. The main objective of `robspack` is to
integrate different CH4 ObsPack text files in a long 
format, adding fields for specific metadata information,
with a tidy structure [@Silge2016]. Hence, the data 
can easily be processed with `data-table`, allowing
to produce plots with `ggplot2` and other R packages [@g2].

The steps followed by `robspack` are described in the 
following lines. First, it is necessary to create a 
summary of the data , read all the file names, identify 
key-words, sectors and full paths. The data is read
iterating on each file for each category such as aircraft, tower, 
surface, flasks, ships and aircore. It is necessary to
add UTC, local time and calculate altitude above ground 
level (agl) from the data. For continuous ground sites, 
afternoon observations are selected and averaged to reduce
correlations and noise. Given the high frequency of 
observations of data, usually every second, aircraft data is
usually averaged every 20 seconds. Once the data is processed
for each category, it is merged in a long format. The 
installation and functions created to perform these tasks 
are shown in the following sections.

# Functions

The `robspack` functions are shown in the table 1.

Table 1. Functions and classes in `robspack`.

| Function              | Description                                       |
|-----------------------|---------------------------------------------------|
| `invfile`             | Class with `print`, `summary` and `plot` methods  |
| `obs_addltime()`      | Add local time based on metadata and longitude    |
| `obs_addtime()`       | Add UTC time                                      |
| `obs_addzero()`       | Format numbers to match specific requirements     |
| `obs_agg()`           | Aggregates ObsPack by time                        |
| `obs_find_receptors()`| Find expected receptors and NetCDF files          |
| `obs_format()`        | Format for some columns of data.table             |
| `obs_freq()`          | Return numeric vector in intervals                |
| `obs_invfiles()`      | Construct `invfile` objects                       |
| `obs_list.dt()`       | Rbind list of data.frames with different names    |
| `obs_meta()`          | Reads ObsPack metadata                            |
| `obs_out()`           | Outersect, opposed as intersect                   |
| `obs_rbind()`         | Rbind data.frames with different names            |
| `obs_read()`          | Read files, and add metadata as columns           |
| `obs_read_csvy()`     | Read csvy file and prints yaml header             |
| `obs_roundtime()`     | Round seconds from "POSIXct" "POSIXt" classes     |
| `obs_summary()`       | Construct summary of ObsPack as a data.frame      |
| `obs_table()`         | Return a data.frame with summary of data          |
| `obs_trunc()`         | Trunc numbers with a desired number of decimals   |
| `obs_write()`         | Write CSVY to disk, YAML followed by tabulated    |

# Installation

To install `robspack`, the user must have installed the
R package `remotes` and run the following script. 
This process will install all the required 
dependencies, such as `data.table`, `cptcity`, an R package with
more than 7000 color palettes, and `lubridate`, a package to manage 
time and dates [@cpt;@lu].
Then, we call the libraries to load the
function into the environment.

```{r, eval = F}
remotes::install_github("ibarraespinosa/robspack")
library(robspack)
library(data.table)
```

```{r, include = F, message=F, warning=F}
library(robspack)
library(data.table)
```

For this manuscript, we are presenting
the application of `robspack` on tower observations.

@@

categories <- c(
  "aircraft-pfp",
  "aircraft-insitu",
  "surface-insitu",
  "aircore",
  "surface-pfp",
  "tower-insitu",
  "shipboard-insitu",
  "flask"
)
obs <- "../../../obspack_ch4_1_GLOBALVIEWplus_v4.0_2021-10-14/data/txt"
index <- obs_summary(obs = obs)

@@

```{r, eval = T}
master <- df5
master$timeUTC <- as.character(master$timeUTC)
master$timeUTC_end <- as.character(master$timeUTC_end)
master$local_time <- as.character(master$local_time)

fwrite(master, 
       file = "tower_insitu_2020.txt",
       sep = " ")
```

The format Comma Separated Value with YAML (CSVY)[^3]
consists in a typical CSV with a YAML header.
The function`obs_write` includes the argument
`notes` which allows adding custom notes at the header of the file. Below the notes, `obs_write` adds 
the output of the R function `str`, which
provides a vertical summary of the data, known as
structure.

```{r, eval = T}
obs_write_csvy(dt = master,
              notes = "tower 2020",
              out = "tower_insitu_2020.csvy")
```

@@

```{r, eval = F}
readLines("tower_insitu_2020.csvy")[1:38]
```

@@

```
## Recommendation for other applications
The approach to generate receptors depends on each type of 
observation and other considerations. For instance, aircraft with 
continuous observations at each second can be filtered and averaged 
every 20 seconds. In that way, the footprints are still 
representative and it would not be necessary to run HYSPLIT 
every second. Of course, it depends on the application 
and objective of the study. For this manuscript, we are 
presenting the generation of receptors based on tower observations. 
Furthermore, in this package we are sharing scripts to process other sectors
The scripts are available in the path 
`https://github.com/ibarraespinosa/robspack/tree/main/rscripts`

## Conclusion

In this manuscript we presented an `robskpack`,
an R package to read and process
$CH_4$ ObsPack GLOBALVIEW+ published by the Global Monitoring 
Laboratory (GML) from the National Oceanographic and Atmospheric
Administration (NOAA). `robspack` reads the text data
which have different headers and organizes them in a common 
format. Then, this software applies calculations to
filter observations by time and space. Finally,
this software generates receptors in a suitable format that 
allows it to run HYSPLIT and generate footprints. 
This software does 
not provide methods to run HYSPLIT, but the user
can follow the site https://www.ready.noaa.gov/HYSPLIT.php.

[^1]: https://h2oai.github.io/db-benchmark/
[^2]: https://gml.noaa.gov/ccgg/carbontracker-lagrange/
[^3]: https://csvy.org/

# Acknowledgements

Funding: This project is funded by the NOAA Climate Program Office AC4
and COM programs (NA21OAR4310233 / NA21OAR4310234). This research was 
supported by the NOAA cooperative agreement NA22OAR4320151.

```{r, include=FALSE}
file.copy(from = "paper.md", to = "../", overwrite = TRUE)
file.copy(from = "paper.bib", to = "../", overwrite = TRUE)
```

# References
